\input{header.tex}
\begin{document}

\title{Permutations as Genome Rearrangements}
\author{Simon Gr\"uning}
\address[Simon Gr\"uning]{University of Zurich, R\"{a}mistrasse 71, 8006 Zurich}
\email[Simon Gr\"uning]{\href{mailto:simon.gruening@uzh.ch}{simon.gruening@uzh.ch}}

\maketitle

\clearpage



\begin{remark}
RECALL: inversion
\end{remark}


\section{Block Interchanges}

\subsection{The Cycle Graph and Its Relation to $bid$}

\begin{definition}
A Block Interchange is an operation that interchanges two blocks of consecutive entries without rearranging said entries.
\end{definition}

\begin{example}
 $ |34|17|562| \rightarrow |562|17|34| $
\end{example}

\begin{definition}
Let the Block Interchange Distance between two n-permutation $p,q$, denoted as $bid(p,q)$, be the smallest number of block interchanges required to transform $p$ into $q$. Let $bid(p) := bid(p,id)$ be the number of interchanges required to sort $p$.
\end{definition}

\begin{remark}
$bid(p,q)$ is a metric and is left-invariant, just as we have seen for $btd(p,q)$.
\end{remark}

\begin{definition}
Let $p = p_1 p_2 \cdots p_n$ be an n-permutation. The  Cycle Graph $G(p)$ is a graph of coloured directed edges on the vertex set $\{ 0,1, \cdots ,n,n+1 \}$ constructed as follows:
\begin{enumerate}
\item For every $i$ with $1 \leq i \leq n+1$: Add a black edge from $p_i$ to $p_{i-1}$
\item For every $i$ with $0 \leq i \leq n$: Add a red edge from $i$ to $i+1$
\end{enumerate}
with $p_0 = 0$ and $p_{n+1} = n+1$.
\end{definition}

\begin{remark}
The Cycle Graph has $2n+2$ edges.
\end{remark}

\begin{example}

$p_1 = 123 \implies c(G(p_1))=4$ 

\[  \xygraph{
!{<0cm,0cm>;<1.5cm,0cm>:<0cm,1.5cm>::}
!{(0,0) }*+{\bullet_{0}}="0"
!{(1,0) }*+{\bullet_{1}}="1"
!{(2,0) }*+{\bullet_{2}}="2"
!{(3,0) }*+{\bullet_{3}}="3"
!{(4,0) }*+{\bullet_{4}}="4"
"4":"3"
"3":"2"
"2":"1"
"1":"0"
"0":@[red]@/^0.5cm/"1"
"1":@[red]@/^0.5cm/"2"
"2":@[red]@/^0.5cm/"3"
"3":@[red]@/^0.5cm/"4"
}  \]

$p_2 = 4213 \implies c(G(p_2))=1$ 

\[  \xygraph{
!{<0cm,0cm>;<1.5cm,0cm>:<0cm,1.5cm>::}
!{(0,0) }*+{\bullet_{0}}="0"
!{(1,0) }*+{\bullet_{4}}="4"
!{(2,0) }*+{\bullet_{2}}="2"
!{(3,0) }*+{\bullet_{1}}="1"
!{(4,0) }*+{\bullet_{3}}="3"
!{(5,0) }*+{\bullet_{5}}="5"
"5":"3"
"3":"1"
"1":"2"
"2":"4"
"4":"0"
"0":@[red]@/^0.5cm/"1"
"1":@[red]@/^0.5cm/"2"
"2":@[red]@/^0.5cm/"3"
"3":@[red]@/^1cm/"4"
"4":@[red]@/^1cm/"5"
}  \]

$p_3 = 4312 \implies c(G(p_3))=3$ 

\[  \xygraph{
!{<0cm,0cm>;<1.5cm,0cm>:<0cm,1.5cm>::}
!{(0,0) }*+{\bullet_{0}}="0"
!{(1,0) }*+{\bullet_{4}}="4"
!{(2,0) }*+{\bullet_{3}}="3"
!{(3,0) }*+{\bullet_{1}}="1"
!{(4,0) }*+{\bullet_{2}}="2"
!{(5,0) }*+{\bullet_{5}}="5"
"5":"2"
"2":"1"
"1":"3"
"3":"4"
"4":"0"
"0":@[red]@/^0.5cm/"1"
"1":@[red]@/^0.5cm/"2"
"2":@[red]@/^0.5cm/"3"
"3":@[red]@/^0.5cm/"4"
"4":@[red]@/^1cm/"5"
}  \]

For the final permutation $p_3$ notice that $bid(p_3)=1=\frac{4+1-3}{2} $, something which will become relevant in a later theorem.
\end{example}

\begin{lemma}
The cycle graph $G(p)$ has a unique decomposition into edge-disjoint directed cycles in which the colors of the edges alternate.
\end{lemma}

\begin{proof}
It is easy to see that a path of alternating colours cannot branch since by construction there is always only one edge per color directed away from every node. We begin at the $0$-node and complete a cycle (this must occur since $G(p)$ is finite). Once we have done so, we remove the edges of this cycle from $G(p)$ and iterate. Notice that in removing full cycles the structure is maintained as we only remove disjointly coloured pairs of edges from every node. The cycles are then edge-disjoint by construction.
\end{proof}

\begin{definition}
Let $c(G(p))$ be the number of alternating directed cycles in the decomposition of $G(p)$. To avoid confusion, let $c(\Gamma (p))$ denote the number of cycles in the traditional sense of a permutation. From now on the cycles of $G(p)$ will refer to the alternating cycles.
\end{definition}

\begin{remark}

\begin{enumerate}
\item The cycles are only edge-disjoint not vertex-disjoint.
\item If there is an alternating path from one node to another, both nodes are in the same cycle.
\item The identity permutation has $n+1$ cycles. In fact, it is the only permutation with this many cycles.
\end{enumerate}
\end{remark}

Before we can find a formula for $bid(p)$ we shall examine how single block interchanges can influence the block interchange distance of a permutation in the following two lemmas.

\begin{definition} Let $p = p_1 p_2 \cdots p_n $ with $ p \neq id $, or in other words, $p$ contains at least one inversion. Select the inversion in $p$ ($ i,j \in [ n ] : p_i > p_j \wedge i < j$) such that $p_j$ is minimal and if necessary, as a second criterion that $p_i$ is maximal. 

Let $x := p_j , y := p_i$ to ease our eyes. By our selection criterion we find that $x-1$ occurs before $y$ in $p$, whilst $y+1$ occurs after $x$. (In the case of $x =1$ simply let $x-1=0$ )
\begin{align*}
p = \cdots (x-1) | \cdots y | \cdots | x \cdots | (y+1) \cdots
\end{align*}

The Canonical Block Interchange is defined as the interchange transforming $p$ into
\begin{align*}
p' = \cdots (x-1) | x \cdots | \cdots | \cdots y | (y+1) \cdots
\end{align*}

\end{definition}

\begin{lemma}
9.10. 
Let $p \in S_n$ with $p \neq id$. Then there exists a block interchange which increases $c(G(p))$ by two.
\end{lemma}

\begin{proof}
We shall prove that the canonical block interchange suffices, or symbolically:
\begin{align*}
 c(G(p')) = c(G(p)) + 2.
\end{align*}

Notice first that any interchange does not change any red edges in $G(p)$. The canonical interchange changes 3 black edges when $x$ and $y$ are adjacent, and 4 edges otherwise. We shall perform a proof by cases.

\begin{enumerate}

\item \underline{$x,y $ adjacent:} The cycle decomposition of $G(p)$ will contain an alternating path through the following nodes in respective order:
\begin{align*}
a \to (x-1) \to x \to y \to (y+1) \to b
\end{align*} 
where $a$ is the entry in $p$ directly to the right of $(x-1)$ and $b$ the one to the left of $(y+1)$.
\begin{align*}
p = \cdots (x-1) | a \cdots y | \cdots | x \cdots b | (y+1) \cdots
\end{align*}
Since a path exists, we find that the entries are part of the same cycle $C$. After performing the canonical interchange, $C$ is split into 3 cycles. In the black edges we now have $x \to (x-1)$, $(y+1) \to y$, and $a \to b$ while the red have been untouched as seen in the following illustration:

$p$

\[  \xygraph{
!{<0cm,0cm>;<1.5cm,0cm>:<0cm,1.5cm>::}
!{(0,0) }*+{\bullet_{(x-1)}}="(x-1)"
!{(1,0) }*+{\bullet_{a}}="a"
!{(2,0) }*+{\bullet_{y}}="y"
!{(3,0) }*+{\bullet_{x}}="x"
!{(4,0) }*+{\bullet_{b}}="b"
!{(5,0) }*+{\bullet_{(y+1)}}="(y+1)"
"(y+1)":"b"
"x":"y"
"a":"(x-1)"
"y":@[red]@/^0.5cm/"(y+1)"
"(x-1)":@[red]@/^0.5cm/"x"
"b":@[blue]@/^0.5cm/"a"
}  \]

$\to p'$

\[  \xygraph{
!{<0cm,0cm>;<1.5cm,0cm>:<0cm,1.5cm>::}
!{(0,0) }*+{\bullet_{(x-1)}}="(x-1)"
!{(1,0) }*+{\bullet_{x}}="x"
!{(2,0) }*+{\bullet_{b}}="b"
!{(3,0) }*+{\bullet_{a}}="a"
!{(4,0) }*+{\bullet_{y}}="y"
!{(5,0) }*+{\bullet_{(y+1)}}="(y+1)"
"(y+1)":"y"
"x":"(x-1)"
"a":"b"
"y":@[red]@/^0.5cm/"(y+1)"
"(x-1)":@[red]@/^0.5cm/"x"
"b":@[blue]@/^0.5cm/"a"
}  \]

Notice the arrow in blue is not necessarily a singular edge, however it is a well defined alternating path.

\item \underline{$x,y $ not adjacent:} We must introduce another notation in this case. Let $a,b,c,d$ be such that they are adjacent to $(x-1), y, x, (y+1)$ respectively pair-wise. Since they are neighbours we will find the alternating path
\begin{align*}
P_a = a \to (x-1) \to x \to c
\end{align*}
and
\begin{align*}
P_b = b \to y \to y+1 \to d
\end{align*}
We must make a further division of cases at this point.

\begin{enumerate}
\item \underline{$P_a, P_b $ are of one cycle:} Denote the cycle as $C_{ab}$. In this case there exists an alternating path $A_{da}$ from $d$ to $a$ with the final edge being grey. See the following illustration:

$p$

\[  \xygraph{
!{<0cm,0cm>;<1.5cm,0cm>:<0cm,1.5cm>::}
!{(0,0) }*+{\bullet_{(x-1)}}="(x-1)"
!{(1,0) }*+{\bullet_{a}}="a"
!{(2,0) }*+{\bullet_{y}}="y"
!{(3,0) }*+{\bullet_{b}}="b"
!{(4,0) }*+{\bullet_{c}}="c"
!{(5,0) }*+{\bullet_{x}}="x"
!{(6,0) }*+{\bullet_{d}}="d"
!{(7,0) }*+{\bullet_{(y+1)}}="(y+1)"
"(y+1)":"d"
"x":"c"
"b":"y"
"a":"(x-1)"
"y":@[red]@/^0.5cm/"(y+1)"
"(x-1)":@[red]@/^0.5cm/"x"
"d":@[blue]@/^0.5cm/"a"
}  \]

$\to p'$

\[  \xygraph{
!{<0cm,0cm>;<1.5cm,0cm>:<0cm,1.5cm>::}
!{(0,0) }*+{\bullet_{(x-1)}}="(x-1)"
!{(1,0) }*+{\bullet_{x}}="x"
!{(2,0) }*+{\bullet_{d}}="d"
!{(3,0) }*+{\bullet_{b}}="b"
!{(4,0) }*+{\bullet_{c}}="c"
!{(5,0) }*+{\bullet_{a}}="a"
!{(6,0) }*+{\bullet_{y}}="y"
!{(7,0) }*+{\bullet_{(y+1)}}="(y+1)"
"(y+1)":"y"
"x":"(x-1)"
"b":"d"
"a":"c"
"y":@[red]@/^0.5cm/"(y+1)"
"(x-1)":@[red]@/^0.5cm/"x"
"d":@[blue]@/^0.5cm/"a"
}  \]

where the blue edge is our path $A_{da}$, the red edges are kept static, and once more we have split one cycle into three, all else remaining the same.

\item \underline{$P_a, P_b $ are of distinct cycles:} Denote these cycles $C_a$ and $C_b$ respectively. We find that each cycle shall split into two, we shall cement this only with a short illustration as the argument is analogue.

$p$

\[  \xygraph{
!{<0cm,0cm>;<1.5cm,0cm>:<0cm,1.5cm>::}
!{(0,0) }*+{\bullet_{(x-1)}}="(x-1)"
!{(1,0) }*+{\bullet_{a}}="a"
!{(2,0) }*+{\bullet_{y}}="y"
!{(3,0) }*+{\bullet_{b}}="b"
!{(4,0) }*+{\bullet_{c}}="c"
!{(5,0) }*+{\bullet_{x}}="x"
!{(6,0) }*+{\bullet_{d}}="d"
!{(7,0) }*+{\bullet_{(y+1)}}="(y+1)"
"(y+1)":"d"
"x":"c"
"b":"y"
"a":"(x-1)"
"y":@[red]@/^0.5cm/"(y+1)"
"(x-1)":@[red]@/^0.5cm/"x"
"d":@[blue]@/^0.5cm/"b"
"c":@[blue]@/^0.5cm/"a"
}  \]

$\to p'$

\[  \xygraph{
!{<0cm,0cm>;<1.5cm,0cm>:<0cm,1.5cm>::}
!{(0,0) }*+{\bullet_{(x-1)}}="(x-1)"
!{(1,0) }*+{\bullet_{x}}="x"
!{(2,0) }*+{\bullet_{d}}="d"
!{(3,0) }*+{\bullet_{b}}="b"
!{(4,0) }*+{\bullet_{c}}="c"
!{(5,0) }*+{\bullet_{a}}="a"
!{(6,0) }*+{\bullet_{y}}="y"
!{(7,0) }*+{\bullet_{(y+1)}}="(y+1)"
"(y+1)":"y"
"x":"(x-1)"
"b":"d"
"a":"c"
"y":@[red]@/^0.5cm/"(y+1)"
"(x-1)":@[red]@/^0.5cm/"x"
"d":@[blue]@/^0.5cm/"b"
"c":@[blue]@/^0.5cm/"a"
}  \]

Where the blue edges are our paths $C_a$ and $C_b$ exempted $(x-1) \to x$ and $y \to (y+1)$.

\end{enumerate}
\end{enumerate}

Having exhausted all cases, we conclude the proof.

\end{proof}

\begin{lemma}
9.11
Let $p \in S_n $. A block interchange cannot increase $c(G(p))$ by more than two.
\end{lemma}

\begin{proof}
We shall only provide a quick sketch of the proof.
?????????????????????????
\end{proof}


\begin{theorem}
9.9
Let $p \in S_n$. Then $bid(p) = \frac{n+1-c(G(p))}{2}$
\end{theorem}

\begin{proof}
By the previous two lemmata we have for any $p \in S_n$ that
\begin{align*}
\frac{n+1-c(G(p))}{2} \leq bid(p) \leq \frac{n+1-c(G(p))}{2}
\end{align*}
since we require a change in $c(G(p))$ of magnitude $( n+1)-c(G(p))$ to achieve 
\begin{align*}
c(G(\tilde{p}))=n+1 \iff \tilde{p} = id
\end{align*}
 and this can be achieved by half the amount of block interchanges at most and at least, thus culminating our desired equality.
\end{proof}

\begin{remark}
We have discovered that we need at most $\left \lfloor{\frac{n}{2}}\right \rfloor $ block interchanges to sort an n-permutation. Furthermore $bid(p) \in \mathbb{N}$ implies that $(n+1)$ and $c(G(p))$ must either both be odd or even.
\end{remark}


\subsection{Average Number of Block Interchanges Required to Sort $p$}

We would now like to find the average number of block interchanges required for any n-permutation. We wish to count the cardinality of the set of n-permutations with a specific number of cycles in order to calculate this average.

\begin{definition}
Define the Hultman Number as
\begin{align*}
S_H(n,k) := | \{ \pi \in S_n : c(G(p)) = k \} |
\end{align*}
\end{definition}

\begin{example}
$S_H(3,4) = 1$ and $S_H(3,2)=5$, thus all six 5 non-trivial 3-permutations have 2 cycles.

Have a glance at the graph of $p = 321$:

\[  \xygraph{
!{<0cm,0cm>;<1.5cm,0cm>:<0cm,1.5cm>::}
!{(0,0) }*+{\bullet_{0}}="0"
!{(1,0) }*+{\bullet_{3}}="3"
!{(2,0) }*+{\bullet_{2}}="2"
!{(3,0) }*+{\bullet_{1}}="1"
!{(4,0) }*+{\bullet_{4}}="4"
"4":"1"
"3":"0"
"2":"3"
"1":"2"
"0":@[red]@/^0.5cm/"1"
"1":@[red]@/^0.5cm/"2"
"2":@[red]@/^0.5cm/"3"
"3":@[red]@/^0.5cm/"4"
}  \]

\end{example}

\begin{theorem}
(9.14) 
\begin{align*}
S_H(n,k) = | \{ (q,r) \in S_n^2 : c(\Gamma (r)) = k \wedge qr = (12 \cdots n(n+1)) \} |
\end{align*}
\end{theorem}

The previous theorem illustrates the connection between $c(G(p))$ and $c(\Gamma (p))$. It allows us to translate our question into a previously explored one. After some complex and difficult papers alleged to require things such as symmetric functions, character theory, and non-elementary integration, we find our answer:

\begin{theorem}
(9.23) The average number of block interchanges needed to sort an n-permutation is
\begin{align*}
\frac{1}{2} \left( n- \frac{1}{ \left \lfloor{\frac{n+2}{2}}\right \rfloor} - \sum^n_{i=2} \frac{1}{i} \right)
\end{align*}\end{theorem}

\begin{appendix}


\end{appendix}

\end{document}